# PART A Univariate Statistics

### Check for Normality

#### Streptococcus bacteria

```{r}
d <- read.table("RTestData.txt", header=TRUE)

plot(table(d$strep), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$strep, pch = 1, frame = FALSE)
qqline(d$strep, col = "steelblue", lwd = 2)
summary(d$strep)
```
#### Sqrt Tranformation Completed
```{r}
d$strep = sqrt(d$strep)
plot(table(d$strep), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$strep, pch = 1, frame = FALSE)
qqline(d$strep, col = "steelblue", lwd = 2)

```

####  Leptotrichia bacteria

```{r}
d <- read.table("RTestData.txt", header=TRUE)

plot(table(d$lepto), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$lepto, pch = 1, frame = FALSE)
qqline(d$lepto, col = "steelblue", lwd = 2)
summary(d$lepto)
```
#### Sqrt Tranformation Completed
```{r}
d$lepto = sqrt(d$lepto)
plot(table(d$lepto), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$lepto, pch = 1, frame = FALSE)
qqline(d$lepto, col = "steelblue", lwd = 2)

```

####  Prevotella bacteria

```{r}
d <- read.table("RTestData.txt", header=TRUE)

plot(table(d$prev), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$prev, pch = 1, frame = FALSE)
qqline(d$prev, col = "steelblue", lwd = 2)
summary(d$prev)
```
#### Sqrt Tranformation Completed
```{r}
d$prev = sqrt(d$prev)
plot(table(d$prev), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$prev, pch = 1, frame = FALSE)
qqline(d$prev, col = "steelblue", lwd = 2)

```

####  Fusobacteria bacteria

```{r}
d <- read.table("RTestData.txt", header=TRUE)

plot(table(d$fuso), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$fuso, pch = 1, frame = FALSE)
qqline(d$fuso, col = "steelblue", lwd = 2)
summary(d$fuso)
```
#### Sqrt Tranformation Completed
```{r}
d$fuso = sqrt(d$fuso)
plot(table(d$fuso), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$fuso, pch = 1, frame = FALSE)
qqline(d$fuso, col = "steelblue", lwd = 2)

```

####  Percentage of Veillonella bacteria

```{r}
d <- read.table("RTestData.txt", header=TRUE)

plot(table(d$veil), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$veil, pch = 1, frame = FALSE)
qqline(d$veil, col = "steelblue", lwd = 2)
summary(d$veil)
```
#### Sqrt Tranformation Completed
```{r}
d$veil = sqrt(d$veil)
plot(table(d$veil), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$veil, pch = 1, frame = FALSE)
qqline(d$veil, col = "steelblue", lwd = 2)

```

####  Depth of the deepest gum pocket in the mouth (in millimeters)

```{r}
d <- read.table("RTestData.txt", header=TRUE)

plot(table(d$deepest), type="h", lwd=4, ylab="Freq", xlab = "Values")
qqnorm(d$deepest, pch = 1, frame = FALSE)
qqline(d$deepest, col = "steelblue", lwd = 2)
summary(d$deepest)
```
#### No Tranformation Needed

### One-way ANOVA/Summary/Boxplots

#### Percentage of Streptococcus bacteria vs Time
``` {r}
one.way <- aov(strep~time, data=d)
summary(one.way)
boxplot(strep~time, data=d)
```

#### Percentage of Leptotrichia bacteria vs Time
``` {r}
one.way <- aov(lepto~time, data=d)
summary(one.way)
boxplot(lepto~time, data=d)
```
### Correlation analysis
#### Strep vs. prev
```{r}

cor.test(d$strep, d$prev)
```
#### Strep vs. fuso

```{r}

cor.test(d$strep, d$fuso)

```

#### Fuso vs. lepto
```{r}

cor.test(d$fuso, d$lepto, data=d)

```

### Linear regressions
#### Strep vs. Deepest
```{r}
d$strep=sqrt(d$strep)
linr <- lm(strep~deepest, data=d)
plot(strep~deepest, data=d)
abline(linr)
```

#### Fuso vs. Deepest
```{r}
d$fuso=sqrt(d$fuso)
linr <- lm(fuso~deepest, data=d)
plot(fuso~deepest, data=d)
abline(linr)
```

# Part B ggplot2

#### Prev Histogram
```{r}
library(ggplot2)
ggplot(d, aes(x=prev)) + geom_histogram(binwidth = 0.5)

```
#### Strep by Deepest Scatterplot
```{r}
ggplot(d, aes(x=strep, y=deepest)) + geom_point(aes(color=factor(time)))
```
#### Fuso by Deepest Scatterplot
```{r}
ggplot(d, aes(x=fuso, y=deepest)) + geom_point(aes(color=factor(time)))

```

#### Smoothed Fuso by Deepest Scatterplot
```{r}
ggplot(d, aes(x=fuso, y=deepest)) + geom_point(aes(color=factor(time))) + stat_smooth(method=lm)
```

# Part C Vegan analysis

```{r}
library(vegan)
d <- read.table("RTestData.txt", header=TRUE)
d.c = d[2:6] #converted from percent to number
d.c
```
#### Alpha Diversity (Shannon)
```{r}
shannonvar <- diversity(d.c, "shannon")
shannonvar
```
#### Rarefy/Plot
```{r}
indvabun <- rowSums(d.c)
indvabun
rarefacmin <- min(rowSums(d.c))
rarefacmin


```
#### Bray-Curtis Dissimilarity Matrix and Histogram
```{r}
brayforrain <- vegdist(d.c, method = "bray")
brayforrain
hist(shannonvar)
```
####Multidimensional Scaling Plot (Bray-Curtis)
```{r}
NMDS.c.d <- metaMDS(comm = d.c, distance = "bray", trace = FALSE, autotransform = FALSE)
plot(NMDS.c.d)
```
####Multidimensional Scaling Plot (Bray-Curtis) Status Colored
```{r}
MSP.xy <- data.frame(NMDS.c.d$points)
MSP.xy$status <- d$status
library(ggplot2)
ggplot(MSP.xy, aes(MDS1, MDS2, color = status)) +
  geom_point() + theme_bw()

```
####Multidimensional Scaling Plot (Bray-Curtis) Time Colored
```{r}
MSP.xy$time <- d$time
ggplot(MSP.xy, aes(MDS1, MDS2, color = time)) +
  geom_point() + theme_bw()

```
